A Realistic Guide to Making Data Available Alongside Code to Improve Reproducibility

### 2019-05-09 paper.Rmd
Fixed more typos

Data are a fundamental currency upon which scientific discoveries are made. Without access to good data, it becomes extremely difficult, if not impossible, to advance science. Yet, a large majority of data on which published research papers are based rarely see the light of day and are only visible to the original authors (citation).  Sharing datasets upon publication of a research paper has benefits to individual researchers, often through increased visibility of the work. But there is also a component of benefit to the broader scientific community. This primarily comes in the form of potential for reuse in other contexts along use for training and teaching (@McKiernan2016). Assuming that the data have no privacy concerns (e.g., human subjects, locations of critically endangered species), or that the act of sharing doesn't put the authors at a competitive disadvantage, sharing data will always have a net positive benefit. First and foremost, sharing data along with other artifacts can make it easier for a reader to independently reproduce the results, thereby increasing transparency and trust in the work. The act of easy data sharing can also improves model training, as many different models can be tried and tested on latest data sources, closing the loop on research and application of statistical techniques. Existing datasets can be combined or linked with new or existing data, fostering the development and synthesis of new ideas and research areas. The biggest of these benefits is the overall increase in reproducibility.

@@

A typical data scientist loads a dozen or two of these open source libraries at the top of a notebook and then rely on existing routines to rapidly read, transform, visualize, and model data. Each of these packages individually depend on a complex web of other packages, building upon existing work rather than reimplementing everything from scratch. Working from script and a list of such dependencies, a data analyst can easily install all the necessary tools in any local or remote environment and reproduce the computation. When new functionality is developed, it is packaged into a separate entity and added to a language's package manager.

@@

To frame discussion around the challenges of data documenting, we can think of two features: "Effort to prepare", and "Ease of understanding". We can plot these on two axes and place datasets along each to indicate the ease of understanding and preparation (see Figure \@ref(fig:fig-effort-underrstanding)).

@@

To improve data impact it needs more time to prepare so it is easier to understand. Accordingly, data with higher potential for impact and reuse should be made as easy to understand as possible. Therefore data with high potential impact requires more time and effort to prepare. For example, data from medical randomized control trials (RCTs) on cancer treatment can have high impact, so requires a lot of time and effort document. Comparatively, a small survey on simple demographics can have low impact. So it might not have a worthwhile tradeoff for ease of understanding and ease of preparation. This means effort put into preparing data documentation for a small survey to be made easy to use should be kept relatively simple and not overcomplicated.

@@

In **Least Moving Parts**, the data might be published with an R package, or as part of a GitHub release using piggyback @piggyback, or a in serverless databaseses. This approach **gets the data somewhere rather than nowhere**. It's minimal features means it is simple to maintain. A downsides is that it does not scale to larger data. Self hosting the data is an option, but we discourage this, as it may succumb to bit rot.

@@

Tooling for producing these this information speeds up the process of sharing data. To help create codebooks, there are R packages such as `dataMeta`, `memisc`. and `codebook`. Codebooks are implemented in other software such as STATA, which provides a "codebook" command. Data can be packaged up in a "data package" with [DataPackageR](https://github.com/RGLab/DataPackageR), which provides tools to wrap up data into an R package, while prioviding helpers with MD5sum checks that track versioning. Note that is different to [Frictionlessdata's tabular data package spec](https://frictionlessdata.io/docs/tabular-data-package/).

@@

4. **Provide a machine readable format** for your data that describes what each column contains. This can be a json schema file, which also helps websites like Google Dataset Search automatically discover your data.
5. **Provide the data in its rawest form** in a folder called "data-raw". This
 ensures provenance of your data by keeping a master copy from which all
other tidied up data can be created.
6. **Provide [open source?] scripts** used to clean data from rawest form into tidy/analysis ready format. These describe the steps taken to prepare the data,
which helps explain and document any decisions taken during the cleaning phase.
These should ideally operate on some raw data stored in the "data-raw" folder (rule 5).
7. **Keep additional copies in more accessible locations**: Even if you archive the data somewhere long-term, keep a second (or additional copies) in locations that are more readily and easily accessible for data analysis workflows. These include services like GitHub, GitHub LFS and others where fast CDNs make access and reuse practical, but fall back to Zenodo or similar in case these services were to go away.
8. **Use a hash function like MD5 checksum** to ensure that the data you deposited are the same that are being made available to others. Hash functions not only remove redundancy by removing the need to download the same data again, but also help ensure that the data are valid.
9. **Only add a citation if your data has a DOI**. A citation only makes sense when your data has a DataCtie compliant DOI, which is automatically provided when data is published in repositories like Zenodo and Dryad. Even then a citation may not accrue references, but without a DOI, it is gauraunteed not to.
10. Store a copy of your data in simple formats such as csv: CSV files follow a flat and extremely simple schema and can be opened by any text editor. These files can also be easily parsed by any data science programming language. Modern software packages allow for fast reads without consuming too much memory and these tools also allow for easy modification of the data.

>>>>

Data are a fundamental currency upon which scientific discoveries are made. Without access to good data, it becomes extremely difficult, if not impossible, to advance science. Yet, a large majority of data on which published research papers are based rarely see the light of day and are only visible to the original authors (citation).  Sharing data sets upon publication of a research paper has benefits to individual researchers, often through increased visibility of the work. But there is also a component of benefit to the broader scientific community. This primarily comes in the form of potential for reuse in other contexts along use for training and teaching (@McKiernan2016). Assuming that the data have no privacy concerns (e.g., human subjects, locations of critically endangered species), or that the act of sharing doesn't put the authors at a competitive disadvantage, sharing data will always have a net positive benefit. First and foremost, sharing data along with other artifacts can make it easier for a reader to independently reproduce the results, thereby increasing transparency and trust in the work. The act of easy data sharing can also improves model training, as many different models can be tried and tested on latest data sources, closing the loop on research and application of statistical techniques. Existing data sets can be combined or linked with new or existing data, fostering the development and synthesis of new ideas and research areas. The biggest of these benefits is the overall increase in reproducibility.

@@

A typical data scientist loads a dozen or two of these open source libraries at the top of a notebook and then rely on existing routines to rapidly read, transform, visualize, and model data. Each of these packages individually depend on a complex web of other packages, building upon existing work rather than re-implementing everything from scratch. Working from script and a list of such dependencies, a data analyst can easily install all the necessary tools in any local or remote environment and reproduce the computation. When new functionality is developed, it is packaged into a separate entity and added to a language's package manager.

@@

To frame discussion around the challenges of data documenting, we can think of two features: "Effort to prepare", and "Ease of understanding". We can plot these on two axes and place datasets along each to indicate the ease of understanding and preparation (see Figure \@ref(fig:fig-effort-understanding)).

@@

To improve data impact it needs more time to prepare so it is easier to understand. Accordingly, data with higher potential for impact and reuse should be made as easy to understand as possible. Therefore data with high potential impact requires more time and effort to prepare. For example, data from medical randomized control trials (RCTs) on cancer treatment can have high impact, so requires a lot of time and effort document. Comparatively, a small survey on simple demographics can have low impact. So it might not have a worthwhile tradeoff for ease of understanding and ease of preparation. This means effort put into preparing data documentation for a small survey to be made easy to use should be kept relatively simple and not over complicated.

@@

In **Least Moving Parts**, the data might be published with an R package, or as part of a GitHub release using piggyback @piggyback, or a in serverless databases. This approach **gets the data somewhere rather than nowhere**. It's minimal features means it is simple to maintain. A downsides is that it does not scale to larger data. Self hosting the data is an option, but we discourage this, as it may succumb to bit rot.

@@

Tooling for producing these this information speeds up the process of sharing data. To help create codebooks, there are R packages such as `dataMeta`, `memisc`. and `codebook`. Codebooks are implemented in other software such as STATA, which provides a "codebook" command. Data can be packaged up in a "data package" with [DataPackageR](https://github.com/RGLab/DataPackageR), which provides tools to wrap up data into an R package, while providing helpers with MD5sum checks that track versioning. Note that is different to [Frictionlessdata's tabular data package spec](https://frictionlessdata.io/docs/tabular-data-package/).

@@

4. **Provide a machine readable format** for your data that describes what each column contains. This can be a json schema file, which also helps websites like Google Dataset Search automatically discover your data.
5. **Provide the data in its rawest form** in a folder called "data-raw". This
 ensures provenance of your data by keeping a master copy from which all
other tidied up data can be created.
6. **Provide [open source?] scripts** used to clean data from rawest form into tidy/analysis ready format. These describe the steps taken to prepare the data,
which helps explain and document any decisions taken during the cleaning phase.
These should ideally operate on some raw data stored in the "data-raw" folder (rule 5).
7. **Keep additional copies in more accessible locations**: Even if you archive the data somewhere long-term, keep a second (or additional copies) in locations that are more readily and easily accessible for data analysis workflows. These include services like GitHub, GitHub LFS and others where fast CDNs make access and reuse practical, but fall back to Zenodo or similar in case these services were to go away.
8. **Use a hash function like MD5 checksum** to ensure that the data you deposited are the same that are being made available to others. Hash functions not only remove redundancy by removing the need to download the same data again, but also help ensure that the data are valid.
9. **Only add a citation if your data has a DOI**. A citation only makes sense when your data has a DataCite compliant DOI, which is automatically provided when data is published in repositories like Zenodo and Dryad. Even then a citation may not accrue references, but without a DOI, it is gauraunteed not to.
10. Store a copy of your data in simple formats such as csv: CSV files follow a flat and extremely simple schema and can be opened by any text editor. These files can also be easily parsed by any data science programming language. Modern software packages allow for fast reads without consuming too much memory and these tools also allow for easy modification of the data.
